Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-11-21 paper/paper.tex
finished LaTex re-formatting

@@

\section{Results}

\subsection{A system to identify and enrich for rare IFN+ cells}

%%% start IFNrare figure %%%
\begin{figure}[b!]
\centerline{
{\bf \Large A}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.02\textwidth}
{\bf \Large B} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
Reporter cells to identify and enrich IFN+ cells.
{\bf (A)} The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
{\bf (B)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (\SUPPFIG{reportervalidation}C).
The plot also shows uninfected cells, and cells infected with Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}
\end{figure}
%%% end IFNrare figure %%%

@@

To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\cite[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}A).
Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

@@

\subsection{Combined transcriptomics and virus-sequencing of single infected cells}

@@

\caption{
Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.
{\bf (A)}
IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
IFN+ cells are enriched by MACS (\SUPPFIG{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
{\bf (B)}
The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
{\bf (C)}
Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
{\bf (D)}
viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are enriched and sequenced by PacBio).
{\bf (E)}
The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
}
\label{fig:workflow}

@@

We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences~\cite[CCSs;][]{travers2010flexible}.
These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).

\subsection{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells}

@@

\caption{
Single-cell transcriptomics of IFN-enriched influenza-infected cells.
{\bf (A)} 
Number of cells for which transcriptomes were obtained.
From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
{\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
{\bf (C)} Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA in the human cells that are called as infected.
{\bf (D)} Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
{\bf (E)} Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
{\bf (F)} Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
{\bf (G)} Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
Many cells that do not express IFN still express ISGs (\SUPPFIG{ISG}).
}
\label{fig:transcriptomics}

@@

\subsection{Full genotypes of viruses infecting single IFN+ and IFN- cells}

%%% start genotypes figure %%%
\begin{figure}
\begin{fullwidth}
{\centering
\includegraphics[height=0.9\textheight]{figures/single_cell_figures/p_genotypes.pdf}
}
\caption{
Viral genotypes and infection outcomes in single cells.
Arrows indicate presence of a viral gene from one (light blue) or both viral barcode variants (dark blue).
Circles and boxes indicate mutations or indels as described in the legend, with circle areas and box heights proportional to the fraction of CCSs with that mutation.
For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively. 
Green boxes at left show the percent of all mRNA in that cell derived from virus.
Orange boxes show the percent of cellular mRNA derived from IFN, with orange box frames indicating cells classified as IFN+ in \FIG{transcriptomics}G.
}
\label{fig:genotypes}

@@

\subsection{Viral defects associated with infection outcome in single cells}

%%% begin mutations figure %%%
\begin{figure}
\begin{fullwidth}
{\centering
\includegraphics[width=\linewidth]{figures/single_cell_figures/p_mutations.pdf}
}
\caption{
Viral features associated with heterogeneity in infection outcome among cells for which we sequenced all expressed viral genes.
{\bf (A)} 
Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight genes.
Cells infected by fully wild-type virus exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
{\bf (B)}
IFN expression among cells expressing unmutated copies of all genes, and among cells with mutations or missing genes.
{\bf (C)}
Specific viral defects associated with IFN induction.
The top panel show the percent of IFN- and IFN+ cells that fail to express each viral gene.
The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each gene, conditioned on the cell expressing that gene.
Numbers give P-values (Fisher's exact test) for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
{\bf (D)}
There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN induction among cells that lack NS.
Throughout this figure, we only consider substitutions that are non-synonymous.
Insertions are not shown as a mutation type as they are extremely rare (\FIG{genotypes}).
}
\label{fig:mutations}

@@

\subsection{Validation that viral defects in single IFN+ cells often cause increased IFN induction}

@@

\centerline{\includegraphics[width=0.85\textwidth]{figures/Validation_Figure/ifn_plot.pdf}}
\caption{Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.
{\bf (A)}
Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line.
The numbers indicate the median of four measurements for each viral mutant.
The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for wild type is indicated with a dashed blue line.
Points are colored orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
{\bf (B)}
Similar to the first panel, but validates the increased IFN induction for a large internal deletion in the PB1 gene.
See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
}
\label{fig:validation}

\figdata{Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\cite{hoffmann2000dna} are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}}
\label{figdata:sequences}

@@

\section{Discussion}
We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
Our method builds on the observation by that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\cite{saikia2018simultaneous, zanini2018virus}.
To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.

@@

\section{Methods and Materials}

\subsection{IFN reporter cell lines}
We created IFN reporter variants of the A549 human lung epithelial cell line (\FIG{IFNrare}A).
The parental A549 cell line used to create these reporters was obtained from ATCC (CCL-185), and was tested as negative for mycoplasma contamination by the Fred Hutch Genomics Core and authenticated using the ATCC STR profiling service.
The cells were maintained in D10 media (DMEM supplemented with 10\% heat-inactivated fetal bovine serum, 2 mM L-glutamine, 100 U of penicillin / ml, and 100 $\mu$g of streptomycin / ml) at 37$^{\circ}$C and 5\% carbon dioxide.

@@

\subsection{Viruses for single-cell experiments}
We performed the single-cell experiments using the A/WSN/1933 (H1N1) strain of influenza virus.
We used both the wild-type virus and a variant of the virus where synonymous mutations were added within a few 100 nucleotides of each termini of each gene segment.
We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\cite{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{CCSs}B).
The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{virus_seqs}.
These genes were cloned into the pHW2000~\cite{hoffmann2000dna} reverse-genetics plasmid.

@@

\subsection{Flow cytometry analyses for HA expression}
For the single-cell experiments (which only examine the transcriptional results of a single cycle of infection), we were most interested in the titer of viral particles that are transcriptionally active for a single round of infection of A549 cells.
We estimated titers of transcriptionally active virions by staining for HA expression in virus-infected A549 cells.
Specifically, we infected A549 cells (or one of the A549 reporter cell line variants as indicated) in influenza growth medium, and at 13 to 14 hours post-infection, we trypsinized cells, re-suspended in phosphate-buffered saline (PBS) supplemented with 2\% heat-inactivated fetal bovine serum (FBS), and stained with 10 $\mu$g/ml of H17-L19, a mouse monoclonal antibody previously shown to bind to the HA from the A/WSN/1933 strain of virus~\cite{doud2017complete}.
After washing in PBS supplemented with 2\% FBS, the cells were stained with a goat anti-mouse IgG antibody conjugated to APC, washed, fixed in 1\% formaldehyde in PBS, washed again, and then analyzed by flow cytometry to determine the fraction expressing detectable HA protein.

\subsection{Single-cell transcriptomics of IFN-enriched infected cells using 10X Chromium}
The single-cell transcriptomics and virus sequencing was performed using the A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-P2A-mNeonGreen reporter.
A schematic of the experiment is shown in \FIG{workflow}.

@@

\subsection{Enrichment and preparation of viral cDNA for PacBio sequencing}
We amplified virus-derived molecules from the small amount of cDNA retained from the 10X Genomics protocol for PacBio sequencing of the full-length cDNA.
All molecules in this cDNA have at their 3' end the cell barcode and UMI plus the constant adaptor sequence that is added during the 10X protocol (see \FIG{workflow} for simple schematic, or the detailed analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or \SUPPFILE{pacbio_analysis} for at detailed schematic).
However, we only wanted to PacBio sequence cDNA molecules derived from virus, since it would be prohibitively expensive to sequence all molecules.
We therefore needed to enrich for the viral molecules, a process made challenging by the need to also retain the 10X adaptor / UMI / cell barcode at the 3' end (the primer at the 5' end can be specific, but the one at the 3' end must bind the common 10X adaptor shared among all molecules).

@@

\subsection{qPCR for viral genes and IFN}
We performed qPCR on reverse-transcribed mRNA for influenza HA (to quantify viral transcription), IFNB1 (to quantify IFN induction), and L32 (a cellular housekeeping gene for normalization).
For the qPCR, we used the SYBR Green PCR Master Mix (Thermo Fisher) according to the manufacturer's protocol using oligo-dT primers.
The qPCR primers were: HA primer 1, 5'-\texttt{GGCCCAACCACACATTCAAC}-3'; HA primer 2, 5'-\texttt{GCTCATCACTGCTAGACGGG}-3'; IFNB1 primer 1, 5'-\texttt{AAACTCATGAGCAGTCTGCA}-3'; IFNB1 primer 2, 5'-\texttt{AGGAGATCTTCAGTTTCGGAGG}-3'; L32 primer 1, 5'-\texttt{AGCTCCCAAAAATAGACGCAC}-3'; L32 primer 2, 5'-\texttt{TTCATAGCAGTAGGCACAAAGG}-3'.

For the qPCR in \SUPPFIG{delqPCR}, A549 cells were seeded at a density of $10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment. 
Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at a MOI of 0.4 based on TCID50 in MDCK-SIAT1 cells.
For the cells with cycloheximide added to block protein expression (and hence secondary transcription), cycloheximide was added to a final concentration of 50 $\mu$g/ml~\cite[a concentration sufficient to block secondary transcription;][]{killip2014activation} at the time of infection.
After 8 hours, mRNA was harvested using the CellAmp Direct RNA Prep Kit for RT-PCR, reverse-transcribed using an oligo-dT primer, and qPCR was performed as described above.

>>>>

% additional packages imported by Jesse Bloom
\usepackage[export]{adjustbox}
\usepackage{float}

% added by Jesse Bloom for full-width figures
\newenvironment{fullwidth}{%
  \begin{adjustwidth}{-2.3in}{}
}{\end{adjustwidth}}

% Added by Jesse Bloom for figures
\renewcommand{\floatpagefraction}{.9}

\newfloat{suppfile}{thp}{lofsupfile}
\floatname{suppfile}{File}
\renewcommand{\thesuppfile}{S\arabic{suppfile}}

\newfloat{suppfig}{thp}{lofsupfile}
\floatname{suppfig}{Fig}
\renewcommand{\thesuppfig}{S\arabic{suppfig}}

\renewcommand{\figureautorefname}{Fig}
\newcommand{\FIG}[1]{\autoref{fig:#1}}

\newcommand{\suppfileautorefname}{File}
\newcommand{\SUPPFILE}[1]{\autoref{suppfile:#1}}

\newcommand{\suppfigautorefname}{Fig}
\newcommand{\SUPPFIG}[1]{\autoref{suppfig:#1}}

@@

\section*{Results}

\subsection*{A system to identify and enrich for rare IFN+ cells}

%%% start IFNrare figure %%%
\begin{figure}[b!]
\centerline{
{\bf \Large A}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.02\textwidth}
{\bf \Large B} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
{\bf Reporter cells to identify and enrich infections that activate IFN expression.}
{\bf (A)} The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
{\bf (B)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (\SUPPFIG{reportervalidation}C).
The plot also shows uninfected cells, and cells infected with Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}
\end{figure}
%%% end IFNrare figure %%%

@@

To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

@@

\subsection*{Combined transcriptomics and virus-sequencing of single infected cells}

@@

\caption{
{\bf Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.}
{\bf (A)}
IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
IFN+ cells are enriched by MACS (\SUPPFIG{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
{\bf (B)}
The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
{\bf (C)}
Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
{\bf (D)}
viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are enriched and sequenced by PacBio).
{\bf (E)}
The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
}
\label{fig:workflow}

@@

We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs~\cite{travers2010flexible}).
These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).

\subsection*{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells}

@@

\caption{
{\bf Single-cell transcriptomics of IFN-enriched influenza-infected cells.}
{\bf (A)} 
Number of cells for which transcriptomes were obtained.
From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
{\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
{\bf (C)} Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA in the human cells that are called as infected.
{\bf (D)} Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
{\bf (E)} Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
{\bf (F)} Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
{\bf (G)} Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
Many cells that do not express IFN still express ISGs (\SUPPFIG{ISG}).
}
\label{fig:transcriptomics}

@@

\subsection*{Full genotypes of viruses infecting single IFN+ and IFN- cells}

%%% start genotypes figure %%%
\begin{figure}
\begin{fullwidth}
{\centering
\includegraphics[height=0.9\textheight]{figures/single_cell_figures/p_genotypes.pdf}
}
\caption{
{\bf Viral genotypes and infection outcomes in single cells.}
Arrows indicate presence of a viral gene from one (light blue) or both viral barcode variants (dark blue).
Circles and boxes indicate mutations or indels as described in the legend, with circle areas and box heights proportional to the fraction of CCSs with that mutation.
For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively. 
Green boxes at left show the percent of all mRNA in that cell derived from virus.
Orange boxes show the percent of cellular mRNA derived from IFN, with orange box frames indicating cells classified as IFN+ in \FIG{transcriptomics}G.
}
\label{fig:genotypes}

@@

\subsection*{Viral defects associated with infection outcome in single cells}

%%% begin mutations figure %%%
\begin{figure}
\begin{fullwidth}
{\centering
\includegraphics[width=\linewidth]{figures/single_cell_figures/p_mutations.pdf}
}
\caption{
{\bf Viral features associated with heterogeneity in infection outcome among cells for which we sequenced all expressed viral genes.}
{\bf (A)} 
Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight genes.
Cells infected by fully wild-type virus exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
{\bf (B)}
IFN expression among cells expressing unmutated copies of all genes, and among cells with mutations or missing genes.
{\bf (C)}
Specific viral defects associated with IFN induction.
The top panel show the percent of IFN- and IFN+ cells that fail to express each viral gene.
The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each gene, conditioned on the cell expressing that gene.
Numbers give P-values (Fisher's exact test) for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
{\bf (D)}
There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN induction among cells that lack NS.
Throughout this figure, we only consider substitutions that are non-synonymous.
Insertions are not shown as a mutation type as they are extremely rare (\FIG{genotypes}).
}
\label{fig:mutations}

@@

\subsection*{Validation that viral defects in single IFN+ cells often cause increased IFN induction}

@@

\centerline{\includegraphics[width=0.85\textwidth]{figures/Validation_Figure/ifn_plot.pdf}}
\caption{
{\bf Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.}
{\bf (A)}
Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line.
The numbers indicate the median of four measurements for each viral mutant.
The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for wild type is indicated with a dashed blue line.
Points are colored orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
{\bf (B)}
Similar to the first panel, but validates the increased IFN induction for a large internal deletion in the PB1 gene.
See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
}
\label{fig:validation}

@@

\section*{Discussion}
We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
Our method builds on the observation by that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\cite{saikia2018simultaneous, zanini2018virus}.
To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.

@@

\section*{Methods and Materials}

\subsection*{IFN reporter cell lines}
We created IFN reporter variants of the A549 human lung epithelial cell line (\FIG{IFNrare}A).
The parental A549 cell line used to create these reporters was obtained from ATCC (CCL-185), and was tested as negative for mycoplasma contamination by the Fred Hutch Genomics Core and authenticated using the ATCC STR profiling service.
The cells were maintained in D10 media (DMEM supplemented with 10\% heat-inactivated fetal bovine serum, 2 mM L-glutamine, 100 U of penicillin / ml, and 100 $\mu$g of streptomycin / ml) at 37$^{\circ}$C and 5\% carbon dioxide.

@@

\subsection*{Viruses for single-cell experiments}
We performed the single-cell experiments using the A/WSN/1933 (H1N1) strain of influenza virus.
We used both the wild-type virus and a variant of the virus where synonymous mutations were added within a few 100 nucleotides of each termini of each gene segment.
We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\cite{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{CCSs}B).
The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{virus_seqs}.
These genes were cloned into the pHW2000~\cite{hoffmann2000dna} reverse-genetics plasmid.

@@

\subsection*{Flow cytometry analyses for HA expression}
For the single-cell experiments (which only examine the transcriptional results of a single cycle of infection), we were most interested in the titer of viral particles that are transcriptionally active for a single round of infection of A549 cells.
We estimated titers of transcriptionally active virions by staining for HA expression in virus-infected A549 cells.
Specifically, we infected A549 cells (or one of the A549 reporter cell line variants as indicated) in influenza growth medium, and at 13 to 14 hours post-infection, we trypsinized cells, re-suspended in phosphate-buffered saline (PBS) supplemented with 2\% heat-inactivated fetal bovine serum (FBS), and stained with 10 $\mu$g/ml of H17-L19, a mouse monoclonal antibody previously shown to bind to the HA from the A/WSN/1933 strain of virus~\cite{doud2017complete}.
After washing in PBS supplemented with 2\% FBS, the cells were stained with a goat anti-mouse IgG antibody conjugated to APC, washed, fixed in 1\% formaldehyde in PBS, washed again, and then analyzed by flow cytometry to determine the fraction expressing detectable HA protein.

\subsection*{Single-cell transcriptomics of IFN-enriched infected cells using 10X Chromium}
The single-cell transcriptomics and virus sequencing was performed using the A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-P2A-mNeonGreen reporter.
A schematic of the experiment is shown in \FIG{workflow}.

@@

\subsection*{Enrichment and preparation of viral cDNA for PacBio sequencing}
We amplified virus-derived molecules from the small amount of cDNA retained from the 10X Genomics protocol for PacBio sequencing of the full-length cDNA.
All molecules in this cDNA have at their 3' end the cell barcode and UMI plus the constant adaptor sequence that is added during the 10X protocol (see \FIG{workflow} for simple schematic, or the detailed analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or \SUPPFILE{pacbio_analysis} for at detailed schematic).
However, we only wanted to PacBio sequence cDNA molecules derived from virus, since it would be prohibitively expensive to sequence all molecules.
We therefore needed to enrich for the viral molecules, a process made challenging by the need to also retain the 10X adaptor / UMI / cell barcode at the 3' end (the primer at the 5' end can be specific, but the one at the 3' end must bind the common 10X adaptor shared among all molecules).

@@

\subsection*{qPCR for viral genes and IFN}
We performed qPCR on reverse-transcribed mRNA for influenza HA (to quantify viral transcription), IFNB1 (to quantify IFN induction), and L32 (a cellular housekeeping gene for normalization).
For the qPCR, we used the SYBR Green PCR Master Mix (Thermo Fisher) according to the manufacturer's protocol using oligo-dT primers.
The qPCR primers were: HA primer 1, 5'-\texttt{GGCCCAACCACACATTCAAC}-3'; HA primer 2, 5'-\texttt{GCTCATCACTGCTAGACGGG}-3'; IFNB1 primer 1, 5'-\texttt{AAACTCATGAGCAGTCTGCA}-3'; IFNB1 primer 2, 5'-\texttt{AGGAGATCTTCAGTTTCGGAGG}-3'; L32 primer 1, 5'-\texttt{AGCTCCCAAAAATAGACGCAC}-3'; L32 primer 2, 5'-\texttt{TTCATAGCAGTAGGCACAAAGG}-3'.

For the qPCR in \SUPPFIG{delqPCR}, A549 cells were seeded at a density of $10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment. 
Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at a MOI of 0.4 based on TCID50 in MDCK-SIAT1 cells.
For the cells with cycloheximide added to block protein expression (and hence secondary transcription), cycloheximide was added to a final concentration of 50 $\mu$g/ml (a concentration sufficient to block secondary transcription~\cite{killip2014activation}) at the time of infection.
After 8 hours, mRNA was harvested using the CellAmp Direct RNA Prep Kit for RT-PCR, reverse-transcribed using an oligo-dT primer, and qPCR was performed as described above.

\subsection*{Viruses and experiments for validation experiments}
In \FIG{validation}, we tested the IFN inducing capacity of a variety of viral mutants identified in the single-cell experiments.
For point-mutant viruses, we created variants for all amino-acid substitutions found in PB1 and NS among IFN+ cells that did not also lack NS.
One of these mutants (amino-acid substitution S704P in PB1) did not reach sufficient titers in a single attempt to generate it by reverse genetics, and so was dropped from the experiment (note that we did not attempt replicates of the reverse genetics for this mutant, and so are \emph{not} confident in drawing strong conclusions about its actual attenuation).
This left six point-mutant viruses: four with point mutations in PB1, and two with point mutations in NS.
We also created a mutant virus that contained the internal deletion in PB1 found in an IFN+ cell.
In addition, we created a virus with an inactivated NS1 to mimic the infections that failed to express NS (we were unable to use complementing cells to generate a viral stock that completely lacked the NS segment).
This NS1stop virus contained six nucleotide changes resulting in the addition of five in-frame stop codons in NS1 starting 10 nucleotides downstream of the 5' splice donor site, thereby disrupting NS1 while leaving NS2 (NEP) intact.
All of these mutants were cloned into the pHW2000 bi-directional reverse-genetics plasmid~\cite{hoffmann2000dna} in order to enable generation of viruses encoding the mutant genes.
 \SUPPFILE{validation_sequences} provides the full sequences for all of these plasmids.
